---
title: "INFX 573 Preliminary Analyses"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preliminary Analyses
```{r}
gtd <- read.csv("globalterrorismdb_0616dist_US_ONLY.csv")
names(gtd)
#install.packages("tidyverse")
library(tidyverse)
```


1) How many unique observations to you have?
```{r}
nrow(gtd)
str(gtd)

vars <- c("iyear", "imonth", "iday", "longitude", "latitude", "city", "provstate", "crit1", "crit2", "crit3", "success", "location", "attacktype1", "suicide", "weaptype1", "weapsubtype1", "targtype1", "targsubtype1", "gname", "ingroup", "nperps", "nkill", "nkillus", "nkillter", "nwound", "nwoundus", "nwoundte", "nperpcap", "property", "propextent", "ihostkid", "hostkid", "doubtterr", "multiple")
```

This dataset includes 2,693 observations of terrorist attacks in the United States.

2) What information/features/characteristics do you have for each observation?
```{r}

```


3) What are the min/max/mean/median/sd values for each of these features?
```{r}
vars <- c("iyear", "imonth", "iday", "longitude", "latitude", "city", "provstate", "crit1", "crit2", "crit3", "success", "location", "attacktype1", "suicide", "weaptype1", "weapsubtype1", "targtype1", "targsubtype1", "gname", "ingroup", "nperps", "nkill", "nkillus", "nkillter", "nwound", "nwoundus", "nwoundte", "nperpcap", "property", "propextent", "ihostkid", "hostkid", "doubtterr", "multiple")


for(var in vars) {
    sum_stats <- summarise(gtd,
    min(var, na.rm = TRUE),
    max(var, na.rm = TRUE),
    mean(var, na.rm = TRUE),
    median(var, na.rm = TRUE),
    sd(var, na.rm = TRUE),
    summary(var, na.rm = TRUE))
}

```


4) What is the distribution of the core features (show a histogram)?
```{r}
```

5) Are there obvious trends in the data (over time, across subgroups, etc.), and are the differences statistically significant?
```{r}
```

6) What are the other salient aspects of the data (e.g. geospatial factors, text content, etc.)
```{r}
```

7) Provide a bullet-list of the next 5-10 tasks you will perform in analyzing your dataset
```{r}

```
